---
  title: "Simple linear regression"
---

The simple linear regression model relates a predictor $x$ to a response $y$ via a linear function with error:

$$
y=\beta_0+\beta_1x+\varepsilon.
$$

This primer leads you down the path of least resistance to fitting this model, creating a scatterplot with the best fit line added to it, and producing a table with the coefficient estimates ($\hat{\beta}_0,\,\hat{\beta}_1$).

## Setup

The commands for working with linear regressions are in the `tidymodels` package, so load that:

```{r}
#| label: load-tidyverse
#| message: false

library(tidyverse) 
library(tidymodels) 
```

Next, we need something to model, so let us load in a data set. We will consider this data set on the stock price of Microsoft and Apple (introduced during Lecture 6 on 9/12/2024):

```{r}
#| label: load-data
#| message: false

stocks <- read_csv("stocks.csv")
```

::: callout-note
Recalling the information [here](https://sta101-f24.github.io/computing/computing-file-paths.html), be prepared to adjust the file path to match how you have organized your files and folders.
:::

To keep things simple, we’ll work with a subset of the data, stock prices in January 2020.

```{r}
stocks_jan2020 <- stocks |>
  filter(month(date) == 1 & year(date) == 2020)
```

## Plotting the least squares regression line ("line of best fit")

This chunk of code creates a scatter plot of the Microsoft and Apple opening stock prices on the various trading days of January 2020, and it adds the line of best fit:

```{r}
ggplot(stocks_jan2020, aes(x = MSFT.Open, y = AAPL.Open)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Daily stock price data",
    subtitle = "January 2020",
    x = "Microsoft opening price",
    y = "Apple opening price"
  )
```

So, all we have to do is add a new `geom_WHAT` layer to a scatterplot to add the line. This is what `geom_smooth` is doing. If you set `se = TRUE`, you get the band indicating the margin of error. Try it out!

## What are the coefficient estimates?

This code will give you a table with the estimates:

```{r}
stock_fit <- linear_reg() |>
  fit(AAPL.Open ~ MSFT.Open, data = stocks_jan2020)

tidy(stock_fit)
```

There is a lot going on in that table, and we will explore some of it later, but focus on the first column for now. This column gives the estimates $\hat{\beta}_0,\,\hat{\beta}_1$. The first row has the estimate $\hat{\beta}_0$ of the intercept, and the second row has the estimate $\hat{\beta}_1$ of the slope. So the fitted model here is: $$
\begin{align*}
\widehat{\text{AAPL}}&=\hat{\beta}_0+\hat{\beta}_1{\text{MSFT}}\\
&\approx3.31+0.45\cdot{\text{MSFT}}.
\end{align*}
$$ So

-   3.31 is the price you would predict for Apple stock if you knew Microsoft stock was opening at \$0;
-   0.45 is the price increase in Apple stock that you would predict if Microsoft stock became more expensive by \$1 (remember, slope = rise/run, so $\Delta\text{AAPL}/\Delta\text{MSFT}$ in this case).

The syntax in the `fit` commands is like `fit(y ~ x)`. So the variable to the left of the `~` will be treated as the response variable ($y$), and the variable to the right will be treated as the predictor ($x$).
