---
  title: "Piping"
---

A *pipe* in `R` is a way of stitching together many commands to make your code easier for a human to read, and to keep it from running off the pages of the PDF documents you will submit to us.

## A very silly example

These two code chunks do exactly the same thing:

```{r}
sum(1, 2)
```

```{r}
1 |>
  sum(2)
```

So the pipe operator `|>` passes (or *pipes*), the number `1` into the `sum` function as the first input. In a simple example like this, `sum(1, 2)` is *definitely* the way I would write the code, but as things get more elaborate, you will want to pipe.

## Tallying stuff up in a spreadsheet

Let's consider the COVID data set from class on 9/3/2024:

```{r}
#| label: load-package-and-data
#| message: false
library(tidyverse)

delta <- read_csv("delta.csv")
```

::: callout-note
Recalling the information [here](https://sta101-f24.github.io/computing/computing-file-paths.html), be prepared to adjust the file path to match how you have organized your files and folders.
:::

A row in this data set is a person, and we record whether or not that person died from/with COVID, and whether or not they were vaccinated. So there are four categories in all:

-   unvaccinated and died
-   unvaccinated and survived
-   vaccinated and died
-   vaccinated and survived

The following code creates a nifty lil' table that tallies up the proportion of people that died versus survived *within each vaccination group*:

```{r}
delta |>
  count(vaccine, outcome) |>
  group_by(vaccine) |>
  mutate(prop = n / sum(n))
```

So within each vaccination group, the numbers sum to one. This code is equivalent, but it provides a truly horrific reading experience:

```{r}
  mutate(group_by(count(delta, vaccine, outcome), vaccine), prop = n / sum(n))
```

## Now you try

The COVID data contains another variable indicating whether or not the person was older or younger than fifty. Write some code (it will be *very* similar to the code above) that produces a table that breaks things down by vaccination status, outcome, *and* age. How many rows should this table contain?
